---
title: "Day 7"
output: html_notebook
---
A giant whale has decided your submarine is its next meal, and it's much faster than you are. There's nowhere to run!

Suddenly, a swarm of crabs (each in its own tiny submarine - it's too deep for them otherwise) zooms in to rescue you! They seem to be preparing to blast a hole in the ocean floor; sensors indicate a massive underground cave system just beyond where they're aiming!

The crab submarines all need to be aligned before they'll have enough power to blast a large enough hole for your submarine to get through. However, it doesn't look like they'll be aligned before the whale catches you! Maybe you can help?

There's one major catch - crab submarines can only move horizontally.

You quickly make a list of the horizontal position of each crab (your puzzle input). Crab submarines have limited fuel, so you need to find a way to make all of their horizontal positions match while requiring them to spend as little fuel as possible.

## Load data

```{r}
library(dplyr)
library(tidyr)
input_file <- "real_input.txt"
crabs <- readr::read_lines(input_file, n_max=1) %>%
  strsplit(",") %>%
  unlist %>%
  as.integer
crabs
```


Determine the horizontal position that the crabs can align to using the least fuel possible. How much fuel must they spend to align to that position?

```{r}
# MAE is minimized by median
final_pos = median(crabs)
final_pos
```

```{r}
fuel_to_reach_median = sum(abs(crabs-final_pos))
fuel_to_reach_median
```

# Part 2

The crabs don't seem interested in your proposed solution. Perhaps you misunderstand crab engineering?

As it turns out, crab submarine engines don't burn fuel at a constant rate. Instead, each change of 1 step in horizontal position costs 1 more unit of fuel than the last: the first step costs 1, the second step costs 2, the third step costs 3, and so on.

As each crab moves, moving further becomes more expensive.


```{r}
# this is MSE, which is minimized by the mean, but we're in integer land so it's fussier
mean_pos = mean(crabs)
possible_means = unique(c(floor(mean_pos), ceiling(mean_pos)))
possible_means
```

```{r}
fuel_used <- function(x) {x*(x+1)/2} # sum_(i=1,x) i
fuel_to_reach_means = purrr::map_dbl(possible_means, ~sum(fuel_used(abs(crabs-.))))
fuel_to_reach_means
```


```{r}
min(fuel_to_reach_means)
```

