---
title: "Day 5"
output: html_notebook
---

You come across a field of hydrothermal vents on the ocean floor! These vents constantly produce large, opaque clouds, so it would be best to avoid them if possible.

They tend to form in lines; the submarine helpfully produces a list of nearby lines of vents (your puzzle input) for you to review.

## Load data

```{r}
library(dplyr)
library(tidyr)
input_file <- "real_input.txt"
segments <- readr::read_lines(input_file) %>%
  tibble(row_text = .) %>%
  mutate(pipe_delim = gsub("(.*),(.*) -> (.*),(.*)","\\1|\\2|\\3|\\4", row_text)) %>%
  separate(pipe_delim, sep="[|]", into=c("x1","y1","x2","y2")) %>%
  select(x1:y2)
segments
```

To avoid the most dangerous areas, you need to determine the number of points where at least two lines overlap. In the above example, this is anywhere in the diagram with a 2 or larger - a total of 5 points.


For now, only consider horizontal and vertical lines: lines where either x1 = x2 or y1 = y2.

```{r}
h_or_v_only <- segments %>%
  filter((x1==x2) | (y1==y2)) %>%
  mutate(idx = row_number()) %>%
  # enforce ordering to simplify intersection logic
  # note: only works on horizontal/vertical lines
  mutate(
    x1_new = pmin(x1, x2),
    x2_new = pmax(x1, x2),
    y1_new = pmin(y1, y2),
    y2_new = pmax(y1, y2),
  ) %>%
  select(idx, ends_with("_new")) %>%
  rename_with(~gsub("_new","",.))
```

Consider only horizontal and vertical lines. At how many points do at least two lines overlap?

```{r}
points_in_line <- function(x1, x2, y1, y2) {
  tibble(x=x1:x2, y=y1:y2)
}
```


```{r}
h_or_v_only %>%
  rowwise %>%
  mutate(points = list(points_in_line(x1, x2, y1, y2))) %>%
  select(points) %>%
  unnest(points) %>%
  count(x,y) %>%
  filter(n>1) %>%
  nrow()
```

# Part 2

Unfortunately, considering only horizontal and vertical lines doesn't give you the full picture; you need to also consider diagonal lines.

Because of the limits of the hydrothermal vent mapping system, the lines in your list will only ever be horizontal, vertical, or a diagonal line at exactly 45 degrees. In other words:

An entry like 1,1 -> 3,3 covers points 1,1, 2,2, and 3,3.
An entry like 9,7 -> 7,9 covers points 9,7, 8,8, and 7,9.


You still need to determine the number of points where at least two lines overlap. 

Consider all of the lines. At how many points do at least two lines overlap?

```{r}
# points_in_line still works on diagonals where length(x1:x2) == length(y1:y2)
segments %>%
  rowwise %>%
  mutate(points = list(points_in_line(x1, x2, y1, y2))) %>%
  select(points) %>%
  unnest(points) %>%
  count(x,y) %>%
  filter(n>1) %>%
  nrow()
```

