---
title: "Day 2"
output: html_notebook
---

the submarine can take a series of commands like forward 1, down 2, or up 3:

* forward X increases the horizontal position by X units.
* down X increases the depth by X units.
* up X decreases the depth by X units.


## Load data

```{r}
library(dplyr)
input = readr::read_delim("input.txt",
                          delim=" ",
                          col_names=c("direction", "count"),
                          col_types = "ci") %>%
  mutate(step = row_number())
input
```


Calculate the horizontal position and depth you would have after following the planned course.

```{r}
total_steps = input %>%
  group_by(direction) %>%
  summarize(count = sum(count)) %>%
  with(structure(as.list(count), names=direction))
total_steps
```

```{r}
final_depth = total_steps$down - total_steps$up
final_depth
```


```{r}
final_horizontal
final_horizontal = total_steps$forward
```

 What do you get if you multiply your final horizontal position by your final depth?
 
```{r}
final_depth * final_horizontal
```
 
 
 
 ## Part 2
 
In addition to horizontal position and depth, you'll also need to track a third value, aim, which also starts at 0. The commands also mean something entirely different than you first thought:

* down X increases your aim by X units.
* up X decreases your aim by X units.
* forward X does two things:
  * It increases your horizontal position by X units.
  * It increases your depth by your aim multiplied by X.
  
Using this new interpretation of the commands, calculate the horizontal position and depth you would have after following the planned course.

```{r}
aim_state <- input %>%
  rename(command = direction) %>%
  mutate(aim_change = case_when(
    command=="up" ~ -count,
    command=="down" ~ count,
    TRUE ~ 0L
  )) %>%
  mutate(aim = cumsum(aim_change))
aim_state
```

```{r}
movements <- aim_state %>%
  filter(command == "forward") %>%
  select(count, aim)
position <- movements %>%
  mutate(horizontal = cumsum(count),
         depth = cumsum(count*aim))
position
```

```{r}
last_position <- tail(position,1)
last_position
```

```{r}
with(last_position, horizontal*depth)
```

