---
title: "Day 15"
output: html_notebook
---

You've almost reached the exit of the cave, but the walls are getting closer together. Your submarine can barely still fit, though; the main problem is that the walls of the cave are covered in chitons, and it would be best not to bump any of them.

The cavern is large, but has a very low ceiling, restricting your motion to two dimensions. The shape of the cavern resembles a square; a quick scan of chiton density produces a map of risk level throughout the cave (your puzzle input). For example:

## Load data

```{r}
library(dplyr)
library(purrr)
library(tidyr)

read <- function(f) {
  lines <- readr::read_lines(f)
  charmat <- strsplit(lines,"") %>%
    do.call(rbind,.)
  mat <- array(0L, dim=dim(charmat))
  mat[] = as.numeric(charmat[])
  mat
}
input <- read("real_input.txt")
demo <- read("demo_input.txt")
demo
```


```{r}
naive_best <- function(risk) {
  # propagate from only down/right
  best <- risk*Inf
  nc <- ncol(risk)
  nr <- nrow(risk)
  d <- dim(risk)
  # start at the lower right (endpoint) and work left, then work up
  best[nr,nc] <- 0
  
  # fill in right column and bottom row
  best[nr,-nc] = rev(cumsum(rev(risk[nr,-1])))
  best[-nr,nc] = rev(cumsum(rev(risk[-1, nc])))
  for(j in rev(seq_len(nc-1))) {
    for(i in rev(seq_len(nr-1))) {
      best[i,j] <- min(
        best[i+1,j]+risk[i+1,j],
        best[i,j+1]+risk[i,j+1]
        )
    }
  }
  best
}

empty_best <- function(risk) {
  best = risk*Inf
  best[nrow(best),ncol(best)] <- 0
  best
}

all_coords <- function(mat) {
  nr=nrow(mat)
  nc=ncol(mat)
  expand_grid(i=rev(seq_len(nr)), j=rev(seq_len(nc))) %>%
    with(map2(i,j,c))
}

lower_right <- function(mat) {
  list(c(nrow(mat),ncol(mat)))
}

optimize_best <- function(risk, best=empty_best(risk), start_check = lower_right(best)) {
  nc <- ncol(risk)
  nr <- nrow(risk)
  d <- dim(risk)
  
  propagate_improvements <- function(pos) {
    cost_to_neighbors <- best[t(pos)]+risk[t(pos)]
    # see who pos can help
    steps = list(r=c(0,1),d=c(-1,0),l=c(0,-1),u=c(1,0))
    check_one <- function(newpos) {
      valid_pos <- all(newpos >= 1, newpos <= d) 
      if(valid_pos) {
        can_improve <- best[t(newpos)] > cost_to_neighbors
        if(can_improve) {
          best[t(newpos)] <<- cost_to_neighbors
          return(list(newpos))
        }
      }
      # either invalid or can't improve
      return(list())
    }
    check_next <- steps %>%
      map(~.+pos) %>%
      map(check_one) %>%
      unlist(recursive=F)
      
    check_next
  }
  
  # iterate until there are no more improvements
  # ensure everything is checked at least once
  check_next <- start_check
  while(length(check_next)>0) {
    this_pos <- check_next[[1]]
    check_later <- check_next[-1]
    new_checks <- propagate_improvements(this_pos)
    check_next <- union(check_later, new_checks)
  }
  best
}
```

```{r}
optimize_best(demo)[1,1]
```

```{r}
optimize_best(input, naive_best(input), start_check = )[1,1]
```

# Part 2

Now that you know how to find low-risk paths in the cave, you can try to find your way out.

The entire cave is actually five times larger in both dimensions than you thought; the area you originally scanned is just one tile in a 5x5 tile area that forms the full map. Your original map tile repeats to the right and downward; each time the tile repeats to the right or downward, all of its risk levels are 1 higher than the tile immediately up or left of it. However, risk levels above 9 wrap back around to 1. So, if your original map had some position with a risk level of 8, then that same position on each of the 25 total tiles would be as follows:

```{r}
tiny <- matrix(1:9,3)
lrep <- function(x, n) map(seq_len(n), function(...) x)
biggify <- function(base) {
  base %>%
  lrep(5) %>%
  map2(as.list(0:4), ~.x+.y) %>%
  do.call(cbind,.) %>%
  lrep(5) %>%
  map2(0:4, ~.x+.y) %>%
  do.call(rbind,.) %>%
  {((.-1)%%9) + 1}
}
big <- biggify(input)
dim(big)
```


```{r}
best_path <- function(best) {
  # start at 1,1 then find the best step until we hit nr,nc
  nr <- nrow(best)
  nc <- ncol(best)
  d <- dim(best)
  stepmat = list(r=c(0,1),d=c(-1,0),l=c(0,-1),u=c(1,0)) %>%
    do.call(cbind,.)
  next_best <- function(pos) {
    next_pos <- stepmat+pos
    valid_next <- ((next_pos >= 1) & (next_pos <= d)) %>%
      colSums %>%
      magrittr::equals(2) %>%
      {next_pos[,.,drop=F]}
    next_vals <- best[t(valid_next)]+risk[t(valid_next)]
    best_next <- valid_next[,which.min(next_vals)]
    c(best_next)
  }
  
  pos=c(1,1)
  steps <- list()
  while(any(pos != c(nr,nc))) {
    pos <- next_best(pos)
    steps <- c(steps, list(pos))
  }
  do.call(rbind, steps)
}
vals <- function(risk) {
  
}
path_points <- function(best, ...) {
  path <- rbind(c(1,1),best_path(best))
  impoints(path[,1], path[,2], ...)
}
```



```{r}
risk <- biggify(input)
naive <- naive_best(risk)
naive_path <- best_path(naive)
mat_to_list <- function(m) map(seq_len(nrow(m)), ~m[.,])
real <- optimize_best(risk, naive, start_check = rev(mat_to_list(naive_path)))
real[1,1]
```


```{r}
disp <- function(x,levs=100,...) {
  col = hcl.colors(levs, "YlOrRd", rev = TRUE)
  breaks = seq(min(x),max(x),length.out=levs+1)
  image(t(x[rev(seq_len(nrow(x))),,drop=F]),col=col,breaks=breaks, ...)
}
impoints <- function(x,y,xmax=max(x), ymax=max(y), ...) {
  points(
    (y-1)/(ymax-1),
    (xmax-x)/(xmax-1),
    ...
  )
}
image_text <- function(mat, ...){
  i = rep(seq(0,1,length.out=nrow(mat)),ncol(mat))
  j = rep(seq(0,1,length.out=ncol(mat)),each=nrow(mat))
  x = j
  y = 1-i
  text(x,y,mat[])
}
disp(demo, useRaster=T, asp=1)
image_text(demo)
```

```{r}
best <- optimize_best(demo)
disp(best,asp=1,useRaster=T, levs=50)
path_points(best, cex=4, pch=19, col="white")
image_text(risk)
```

```{r}
best_path_cost <- function(best, risk) {
  path <- best_path(best)
  step_cost <- risk[path]
  sum(step_cost)
}
best_path_cost(best, demo)
```

